ProteInfer: deep networks for predicting functional properties of proteins

### 2021-09-20 latex/draft3.tex
More precision re: embedding training
Updates

Each model was trained for about 60 hours using the Adam optimizer \citep{kingma2014adam} on 8 NVIDIA P100 GPUs with data parallelism \cite{dean2012large, dataparalellism}.  We found that using more than one GPU for training improved training time by allowing an increased batch size, but did not have a substantial impact on accuracy compared to training for longer with a smaller learning rate and smaller batch size on one GPU. The models have a small set of hyperparameters, such as the number of layers and the number of filters in each layer, which were tuned using random sampling to maximize $\texttt{F}_{max}$ on the random train-test split. Hyperparameter values are available in the Supplement.
\subsection*{A machine-learning compatible dataset for protein function prediction}
The UniProt database is the central global repository for information about proteins. The manually curated portion, Swiss-Prot, is constructed by assessing $~$60,000 papers each year to harvest $\gt$ 35\% of the theoretically curatable information in the literature \cite{uniprotlatest}. We focus on Swiss-Prot to ensure that our models learn from human-curated labels, rather than labels generated by a computational annotation pipeline. Each protein in Swiss-Prot goes through a 6-stage process of sequence curation, sequence analysis, literature curation, family-based curation, evidence attribution, and quality assurance.
Functional annotation is stored in UniProt largely through \textit{database cross-references}, which link a specific protein with a label from a particular ontology. These cross-references include: Enzyme Commission (EC) numbers, representing the function of an enzyme; Gene Ontology (GO) terms relating to the protein's molecular function, biological process, or subcellular localisation; protein family information contained in the Pfam \citep{pfamlatest}, SUPFAM\citep{supfamlatest}, PRINTS \citep{prints}, TIGR \citep{tigr}, PANTHR \citep{panther} databases, or the umbrella database InterPro \citep{interpro}; as well as other information including ortholog databases and references on PubMed. Here, we focus on EC and GO labels, though our model training framework can immediately extend to other label sets.

@@

To facilitate further development of machine learning methods, we provide TensorFlow \citep{abadi2016tensorflow} TFrecord files for Swiss-Prot \footnote{\url{https://console.cloud.google.com/storage/browser/brain-genomics-public/research/proteins/proteinfer/datasets/}}. Each example has three fields: the UniProt accession, the amino acid sequence, and a list of database cross-reference labels. UniProt annotations include only leaf nodes for hierarchical taxononomies such as EC and GO. To allow machine-learning algorithms to model this hierarchy, we added all parental annotations to each leaf node during dataset creation. 
\section*{Results}
\subsection*{Prediction of catalysed reactions}
We initially trained a model to predict enzymatic catalytic activities from amino acid sequence. This data is recorded as Enzyme Commission (EC) numbers, which describe a hierarchy of catalytic functions. For instance, \textbeta{} amylase enzymes have an EC number of EC:3.2.1.2, which represents the leaf node in the following hierarchy:
%%%ECHIERARCHY%%%
%startleaveout
\renewcommand\DTstyle{\rmfamily}
\setlength{\DTbaselineskip}{15pt}
\medskip
\dirtree{%
.1 {\textbf{EC:3.--.--.--} (hydrolases)}.
.2 {\textbf{EC:3.2.--.--} (glycolsylases)}.
.3 {\textbf{EC:3.2.1.--} (glycosidases)}.
.4 {\textbf{EC:3.2.1.2} (hydrolysis of (1$\rightarrow$4)-\textalpha-D-glucosidic linkages in polysaccharides)}.
}
\medskip
%stopleaveout
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{Figures/ec_performance_by_top_level_class.png}%
  \caption{\textbf{ProteInfer performance for predictions within all 7 top-level enzyme groups.}}
  \label{fig:toplevel}
\end{figure}
Individual protein sequences can be annotated with zero (non-enzymatic proteins), one (enzymes with a single function) or many (multi-functional enzymes) leaf-level EC numbers. These are drawn from a total of 8,162 catalogued chemical reactions.
Our best $\texttt{F}_{max}$ was achieved by a model containing 5 residual blocks with 1100 filters each (full details in Supplement). For the dev set, $\texttt{F}_{max}$ converged within 500,000 training steps. On the random split, the model achieves $\texttt{F}_{max}$ = 0.977 (0.976-0.978) on the held-out test data. At the corresponding confidence threshold, the model correctly predicts 96.7\% of true labels, with a false positive rate of 1.4\%. Results from the clustered test set are discussed below.
Performance was roughly similar across labels at the top of the EC hierarchy, with the highest $\texttt{F}_{max}$ score observed for ligases (0.993), and the lowest for oxidoreductases (0.963) (\Cref{fig:toplevel}A%htmlver: 
). For all classes, the precision of the network was higher than the recall at the threshold maximising $\texttt{F}_{max}$. Precision and recall can be traded off against each other by adjusting the confidence threshold at which the network outputs a prediction, creating the curves shown in \Cref{fig:toplevel}B.%htmlver:the figure below
%%%INITIALANALYSIS%%%

@@

We then assessed the ability of this method to more generally localize function within a sequence, even though the model was not trained with any explicit localization information. We selected all enzymes from Swiss-Prot that had two separate leaf-node EC labels for which our model predicted known EC labels, and these labels were mappable to corresponding Pfam labels. For each of these proteins, we obtained coarse-grained functional localization by using CAM to predict the order of the domains in the sequence and compared to the ground truth Pfam domain ordering (see Supplement for details of the method)%htmlver:see Supplement
. We found that in 296 of 304 (97\%) of the cases, we correctly predicted the ordering, though we note that the set of bifunctional enzymes for which this analysis is applicable is limited in its functional diversity (see Supplement)%htmlver:see Supplement 
. Although we did not find that fine-grained, per-residue functional localization arose from our application of CAM, we found that it reliably provided coarse-grained annotation of domains' order, as supported by Pfam. This experiment suggests that this is a promising future area for research.
\subsection*{Neural networks learn a general-purpose embedding space for protein function}
Whereas InterProScan compares each sequence against more than 50,000 %from summing integrated signatures at https://www.ebi.ac.uk/interpro/release_notes.html
individual signatures and BLAST compares against an even larger sequence database, ProteInfer uses a single deep model to extract features from sequences that directly predict protein function. One convenient property of this approach is that in the penultimate layer of the network each protein is expressed as a single point in a high-dimensional space. To investigate to what extent this space is useful in examining enzymatic function, we used the ProteInfer EC model trained on the random split to embed each test set protein sequence into a 1100-dimensional vector.

@@

To investigate this, we took the subset of proteins in Swiss-Prot that are non-enzymes, and so lack any EC annotations. The network would achieve perfect accuracy on these examples if it e.g. mapped all of them to a single embedding that corresponds to zero predicted probability for every enzymatic label. Do these proteins therefore share the same representation in embedding space? The UMAP projection of these sequences' embeddings revealed clear structure to the embedding space, which we visualised by highlighting several GO annotations which the network was never supervised on. For example, one region of the embedding space contained ribosomal proteins, while other regions could be identified containing nucleotide binding proteins, or membrane proteins (\Cref{fig:unsupervised_embedding}%htmlver:interactive figure below
).
%%%nonenzymes%%%
To quantitatively measure whether these embeddings capture the function of non-enzyme proteins, we trained a simple random forest classification model that used these embeddings to predict whether a protein was annotated with the \textit{intrinsic component of membrane} GO term. We trained on a small set of non-enzymes, with ~500 membrane proteins, and evaluated on the rest of the examples. This simple model achieved a precision of 97\% and recall of 60\% for an F1 score of 0.74. Model training and data-labelling took around 15 seconds. This demonstrates the power of embeddings to simplify other studies with limited labeled data, as has been observed in recent work \citep{Alley589333,biswas2020low}.
%%%SWEEP_embedding%%%

@@

An attractive property of deep learning models is that they can be run efficiently, using consumer graphics cards for acceleration. Indeed, recently, a framework has been developed to allow models developed in TensorFlow to be run locally using simply a user's browser \citep{tfjs}, but to our knowledge this has never been deployed to investigate biological sequence data. We therefore built a tool to allow near-instantaneous prediction of protein functional properties in the browser. When the user loads the tool, lightweight EC (5MB) and GO model (7MB) prediction models are downloaded and all predictions are then performed locally, with query sequences never leaving the user's computer. Inference in the browser for a 1500 amino-acid sequence takes < 1.5 seconds for both models (\Cref{sup:js_timing}%htmlver:see Supplement
).
% \begin{table*}[b]
% \centering
% \caption{Bla}
% \begin{tabular}{ccccc}
% \hline
% Input               & Processor & Statistic                                 & InterProScan & ProteInfer  \\ \hline
% $1\times$ 500 aa sequence  & CPU       & Inference time   & 1 min 18 secs            & x           \\ 
% $1\times$ 500 aa sequence  & GPU       & Inference time   & \xmark️          & x           \\ 
% $ (test set) \times$ 500 aa sequences & 12x CPUs       & Inference time & 12 hours, 40 minutes            & x           \\ 
% $64\times$ 500 aa sequence & GPU       & Inference time & \xmark️          & x           \\ 
% --                  & --        & Model sizes (compressed)                  & 50 GB        & 50 MB       \\ 
% --                  & --        & Deliverable within the user’s browser     & \xmark️           &\cmark \\ \hline
% \end{tabular}
% \end{table*}
% \note{
% %I tried making "chimaeric proteins" - when N terminus of amylase is added to MAPK7 that is predicted to have extracellular localisation (otherwise unchanged). With reverse, amylase functionality is unchanged but location changes to intracellular.
% }
\subsection*{Comparison to experimental data}
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{Figures/price.png}%
  \caption{\textbf{ProteInfer predictions for a set of genes recently experimentally reannotated by high-throughput phenotyping.} ProteInfer makes confident and largely accurate predictions at the earliest levels of the EC hierarchy. Accuracy falls at the finest levels of classification (for this set of challenging genes) but fortunately the network declines to make a prediction in most cases, with every label failing to meet the threshold for positive classification. \label{fig:levels}}
\end{figure}
\begin{figure*}[b]
\centering
  \includegraphics[width=\textwidth]{Figures/tetrahydrofolate_synthase.png}
  % See https://docs.google.com/presentation/d/1PqP-Ux7far9NMYvroeZIQQ9iXi41xira7-awDC4w_ss/edit#slide=id.p for the provenance of this figure
  \caption{\textbf{Linking sequence regions to function with class activation mapping for C-1-tetrahydrofolate synthase (accession P11586).} A. Ground truth annotation of function on Uniprot \citep{uniprotlatest}. B. The three horizontal bars are the sequence region ProteInfer predicts are most involved in each corresponding reaction. This concurs with the known function localization. }
  \label{fig:cam}
\end{figure*}

@@

%startleaveout
\section*{Bibliography}
\bibliography{../bibliography}
\onecolumn
\newpage
%stopleaveout
%startleaveout

@@

%stopleaveout
%%%BOTTOMSTUFF%%%
%startleaveout
\section*{Supplement}
\beginsupplement

\subsection{implementation details}

\subsubsection*{Label inheritance}
Some annotations, such as GO terms and EC numbers, come as directed acyclic graphics (which take the form simple trees for EC numbers). Typically in such cases the annotation provided on UniProt is the most-specific that is known. For example if a protein is known to exhibit \textit{sequence specific DNA binding} (GO:0043565). It will not separately be annotated with the ancestral term \textit{DNA binding} (GO:0003677), this is simply assumed from the ontology. Using such an annotations directly, however, is likely to be problematic in a deep learning setting. Failing to annotate an example with the parental term demands that the model predict that the example is negative for this term, which is not the effect we want.
To address this our datasets include labels for all ancestors of applied labels for EC, GO and InterPro datasets. In the case of GO we restrict these to \textit{is\_a} relationships.

\subsubsection*{Class activation mapping}
Many proteins have multiple functional properties. For example we analyse the case of the bifunctional dhfr/ts of \textit{T. gondii}. Such bifunctional enzymes are often not unique -- it is functionally advantageous for these enzymes to be fused, which facilitates channeling of substrate between their active sites. Since there are a number of such examples within Swiss-Prot, the mere existence of a TS domain in a protein is (mild) evidence for possible DHFR function. To increase the interpretability of the network we introduce a technique that simplifies such cases by subtracting the activations for other predicted classes during class activation mapping.

\subsection*{Model architecture}
To create an architecture capable of receiving a wide range of input sequences, with computational requirements determined for each inference by the length of the individual input sequence, we employed a dilated convolutional approach \citep{dilatedconvolutions}. Computation for both training and prediction in such a model can be parallelized across the length of the sequence. By training on full length proteins, in a multi-label training setting, we aimed to build networks that could  extract functional information from raw amino acid sequences. One helpful feature of this architecture is its flexibility with regards to sequence length. Natural protein sequences can vary in length by at least three orders of magnitude, but some architectures have computational requirements that scale with the maximum sequence they are capable of receiving as input, rather than the sequence being currently examined. These fixed-length approaches reduce efficiency as well as place a hard limit on the length of sequences that can be examined.

\subsection{Input data statistics}

We use Swiss-Prot version 2019\_01 in our analysis, which gives us 559077 proteins, or 548264 after filtering for only 20 standard amino acids and filtering fragments.
\begin{table}[htbp]
\centering
\begin{tabular}{|l|r|}
\hline
         Fold &  Number of sequences \\
\Xhline{2pt}
        train &               438522 \\
\hline
          dev &               55453 \\
\hline
         test &               54289 \\
\hline
 all together &               548264 \\
\hline
\end{tabular}
\caption{In our random split of the training data, we allocate about 80\% to the training fold, 10\% to the development fold, and 10\% to the test fold.}
\end{table}
\begin{table}[htbp]
\centering
\begin{tabular}{|l|r|}
\hline
         Fold &  Number of sequences \\
\Xhline{2pt}
        train &               182965 \\
\hline
          dev &               180309 \\
\hline
         test &               183475 \\
\hline
 all together &               546749 \\
\hline
\end{tabular}
\caption{In our clustered split of the training data, we use UniRef50, and allocate approximately equal numbers of sequences to each fold.}
\end{table}
Because different protein functions have differing prevalence, we note the number of proteins that have a given function for Pfam, EC, and GO labels, as well as noting the number of labels per protein.
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/ec_label_distribution.png}
  \caption{Histogram of number of labels per sequence, including hierarchical labels, on the random dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/go_label_distribution.png}
  \caption{Histogram of number of labels per sequence, including hierarchical labels, on the random dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/sequences_per_label_ec_random.png}
  \caption{Number of sequences annotated with a given functional label. (EC class) in the random dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/sequences_per_label_go_random.png}
  \caption{Number of sequences annotated with a given functional label. (GO label) in the random dataset.}
\end{figure}
\pagebreak
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/sequences_per_label_ec_clustered.png}
  \caption{Number of sequences annotated with a given functional label. (EC class) in the clustered dataset.}
\end{figure}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/sequences_per_label_go_clustered.png}
  \caption{Number of sequences annotated with a given functional label. (GO label) in the clustered dataset.}
\end{figure}
\pagebreak
When assigning examples to folds in our clustered dataset, we note that there are test examples that have labels that are never seen in the training data. We report these cases below as ``Impossible'' test example-label pairs.
\begin{table}[htbp]
\centering
\begin{tabular}{|l|r|}
\hline
                                Type &  Number \\
\Xhline{2pt}
                        Train labels &    3411 \\
\hline
                         Test labels &    3414 \\
\hline
              Impossible test labels &    1043 \\
\hline
           Train example-label pairs &  348105 \\
\hline
            Test example-label pairs &  348755 \\
\hline
 Impossible test example-label pairs &    3415 \\
\hline
\end{tabular}
\caption{Clustered dataset statistics for EC labels.}
\end{table}
\begin{table}[htbp]
\centering
\begin{tabular}{|l|r|}
\hline
                                Type &   Number \\
\Xhline{2pt}
                        Train labels &    26538 \\
\hline
                         Test labels &    26666 \\
\hline
              Impossible test labels &     3739 \\
\hline
           Train example-label pairs &  8338584 \\
\hline
            Test example-label pairs &  8424299 \\
\hline
 Impossible test example-label pairs &    11137 \\
\hline
\end{tabular}
\caption{Clustered dataset statistics for GO labels.}
\end{table}
\pagebreak
\subsection{Precision/Recall curves}

\begin{figure}[ht] 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{Figures/random_truncated.pdf}
    \caption{EC number prediction: random split} 
    \label{fig:with_ensemble:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{Figures/random_truncated_go.pdf}
    \caption{GO term prediction: random split} 
    \label{fig:with_ensemble:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{Figures/clustered_truncated.pdf}
    \caption{EC number prediction: clustered split} 
    \label{fig:with_ensemble:c} 
  \end{subfigure}%%
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=1\linewidth]{Figures/clustered_truncated_go.pdf}
    \caption{GO term prediction: clustered split} 
    \label{fig:with_ensemble:d} 
  \end{subfigure} 
  \caption{Bootstrapped precision-recall curves for EC number prediction and gene ontology term prediction for random and clustered splits for four methods: BLAST top pick, single ProteInfer CNN, ensembled ProteInfer CNNs, and ensembled ProteInfer CNNs scaled by BLAST score. }
  \label{fig:with_ensemble} 
\end{figure}

\begin{figure}[ht] 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/prc_ec_random_unzoomed.png}
    \caption{EC number prediction: random split} 
    \label{fig:without_ensemble:a} 
    \vspace{4ex}
  \end{subfigure}%% 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/prc_go_random_unzoomed.png}
    \caption{GO term prediction: random split} 
    \label{fig:without_ensemble:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/prc_ec_clustered_unzoomed.png}
    \caption{EC number prediction: clustered split} 
    \label{fig:without_ensemble:c} 
  \end{subfigure}%%
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[width=0.9\linewidth]{Figures/prc_go_clustered_unzoomed.png}
    \caption{GO term prediction: clustered split} 
    \label{fig:without_ensemble:d} 
  \end{subfigure} 
  \caption{Full precision-recall curves for EC number prediction and gene ontology term prediction for random and clustered splits for four methods: BLAST top pick, single ProteInfer CNN, ensembled ProteInfer CNNs }
  \label{fig:without_ensemble} 
\end{figure}

\clearpage

\subsection{Stratified performance}
\begin{figure}[htbp]
\centering
  \includegraphics[width=\textwidth]{Figures/ec_f1_by_training_size.png}
  \caption{Performance of EC model stratified by number of training examples available for each test example.}
  \label{sup:fig:ec_f1_by_training_size}
\end{figure}

\clearpage

\subsection{Hyperparameters}
We tuned over batch size, dilation rate, filters, first dilated layer, kernel size, learning rate, number of layers, mean vs max pooling, and Adam $\beta_1$, $\beta_2$ and $\epsilon$~\citep{kingma2014adam} over a number of studies to determine the set of parameters that optimized $\texttt{F}_{max}$. We found, as in~\cite{bileschi}, that the network was relatively unresponsive to slight changes in hyperparameters, and found that many of the hyperparameters that performed well in~\cite{bileschi} also performed well for this task. We chose to keep identical hyperparameters for the EC and GO tasks across both the random and clustered splits for simplicity, and we note that parameters with good performance on the random task performed respectively well on the clustered split. 
\begin{table}[htbp]
\centering
\def\arraystretch{1.1}
\begin{tabular}{|c|c|c|c|c|}
\hline
                           & \\
                           & CNN \\
\Xhline{1pt}
concurrent batches (data parallelism) & 8 \\
\hline
batch size                 & \shortstack{40 (per each GPU) \\ Dynamic based on sequence length} \\
\hline
dilation rate              & 3 \\
\hline
filters                    & 1100  \\
\hline
first dilated layer        & 2 \\
% zero-indexed so added one to the values shown in the hparams_registry for clarity.
\hline
gradient clip              & 1 \\
\hline
kernel size                & 9 \\
\hline
learning rate              & 1.5E-3 \\
\hline
learning rate decay rate   & 0.997 \\
\hline
learning rate decay steps  & 1000 \\
\hline
learning rate warmup steps & 3000 \\
\hline
Adam $\beta_1$              & .9 \\
\hline
Adam $\beta_2$              & .999 \\
\hline
Adam $\epsilon$             & 1E-8 \\
\hline
number of ResNet layers     & 5 \\
\hline
pooling                     & mean \\
\hline
ResNet bottleneck factor    & 0.5 \\
\hline
train steps                 & 500000 \\
\hline
\end{tabular}
\caption{Hyperparameters used in convolutional neural networks. We note that hyperparameters for single-GPU training are available in \url{github.com/google-research/proteinfer/blob/master/hparams_sets.py}.}
\label{tab:nn_hparams}
\end{table} 
\pagebreak
\subsection{Predicting coarse-grained functional localization with CAM}
The goal of this experimental methodology is to measure whether or not we correctly order the localization of function in bifunctional enzymes. As such, first we have to identify a set of candidates for experimentation.

\subsubsection{Candidate set construction}
We note that no functional localization information was available to our models during training, so we can consider not just the dev and test sets, but instead the entirety of Swiss-Prot for our experimentation. As such, we take all examples from Swiss-Prot that have an EC label, and convert these labels to Pfam labels using a set of 1515 EC-Pfam manually curated label correlations from InterPro~\cite{mitchell2015interpro}, omitting unmapped labels. We then take the set of 3046 proteins where exactly two of their ground-truth labels map to corresponding Pfam labels. In our Swiss-Prot random test-train split test set, on bifunctional enzymes, we get $0.995$ precision and $0.948$ recall at a threshold of $0.5$, so we believe this set is a reasonable test set for ordering analysis.

We then predict EC labels for these proteins with one of our trained convolutional neural network classifier, considering only the most specific labels in the hierarchy. Then, we map these predicted EC labels to Pfam labels using the InterPro mapping again, and retain only the proteins on which we predict exactly two labels above a threshold of $0.5$, and are left with 2679 proteins. In 2669 out of 2679 proteins, our predictions are identical to the Pfam-mapped ground truth labels.

We take these 2669 that have two true and predicted \textit{Pfam} labels, and look at their current Pfam labels annotated in Swiss-Prot. Of these 2669 proteins, 304 of them contain both of the mapped labels. We note that this difference between 2669 and 304 is likely due in part to Pfam being conservative in calling family members, potential agreements at the Pfam clan vs family level, as well as database version skew issues.

\subsubsection{Computation of domain ordering}
On these 304 proteins we have the same predicted-EC-to-Pfam labels and the same true-Pfam labels. For each of these proteins, we can get an ordering of their two enzymatic domains from Pfam, giving us a \textit{true} ordering. It is now our task to produce a predicted ordering.

We use class activation mapping (CAM) to compute a confidence for each class at each residue for every protein in this set of 304. We then filter this large matrix of values and only consider the families for which our classifier predicted membership, giving us a matrix of shape sequence-length by predicted-classes (which is two in this case). For each class, we take the CAM output and compute a center of mass. Then we order the two classes based on where their center of mass lies.

>>>>

Each model was trained for about 60 hours using the Adam optimizer \citep{kingma2014adam} on 8 NVIDIA P100 GPUs with data parallelism \cite{dean2012large, dataparalellism}.  We found that using more than one GPU for training improved training time by allowing an increased batch size, but did not have a substantial impact on accuracy compared to training for longer with a smaller learning rate and smaller batch size on one GPU. The models have a small set of hyperparameters, such as the number of layers and the number of filters in each layer, which were tuned using random sampling to maximize $\texttt{F}_{max}$ on the random train-test split. Hyperparameter values are available in the \textnormal{ Supplement}%htmlver:<a href="./latex/supplement.pdf">Supplement</a>
.
\subsection*{A machine-learning compatible dataset for protein function prediction}
The UniProt database is the central global repository for information about proteins. The manually curated portion, Swiss-Prot, is constructed by assessing $~$60,000 papers each year to harvest $\gt$ 35\% of the theoretically curatable information in the literature \cite{uniprotlatest}. We focus on Swiss-Prot to ensure that our models learn from human-curated labels, rather than labels generated by a computational annotation pipeline. Each protein in Swiss-Prot goes through a 6-stage process of sequence curation, sequence analysis, literature curation, family-based curation, evidence attribution, and quality assurance.
Functional annotation is stored in UniProt largely through \textit{database cross-references}, which link a specific protein with a label from a particular ontology. These cross-references include: Enzyme Commission (EC) numbers, representing the function of an enzyme; Gene Ontology (GO) terms relating to the protein's molecular function, biological process, or subcellular localisation; protein family information contained in the Pfam \citep{pfamlatest}, SUPFAM\citep{supfamlatest}, PRINTS \citep{prints}, TIGR \citep{tigr}, PANTHR \citep{panther} databases, or the umbrella database InterPro \citep{interpro}; as well as other information including ortholog databases and references on PubMed. Here, we focus on EC and GO labels, though our model training framework can immediately extend to other label sets.

@@

To facilitate further development of machine learning methods, we provide TensorFlow \citep{abadi2016tensorflow} TFrecord files for Swiss-Prot \footnote{\url{https://console.cloud.google.com/storage/browser/brain-genomics-public/research/proteins/proteinfer/datasets/}}. Each example has three fields: the UniProt accession, the amino acid sequence, and a list of database cross-reference labels. UniProt annotations include only leaf nodes for hierarchical taxononomies such as EC and GO. To allow machine-learning algorithms to model this hierarchy, we added all parental annotations to each leaf node during dataset creation. 
\section*{Results}
\subsection*{Prediction of catalysed reactions}
We initially trained a model to predict enzymatic catalytic activities from amino acid sequence. This data is recorded as Enzyme Commission (EC) numbers, which describe a hierarchy of catalytic functions. For instance, \textbeta{} amylase enzymes have an EC number of EC:3.2.1.2, which represents the leaf node in the following hierarchy:
%%%ECHIERARCHY%%%
%startleaveout
\renewcommand\DTstyle{\rmfamily}
\setlength{\DTbaselineskip}{15pt}
\medskip
\dirtree{%
.1 {\textbf{EC:3.--.--.--} (hydrolases)}.
.2 {\textbf{EC:3.2.--.--} (glycolsylases)}.
.3 {\textbf{EC:3.2.1.--} (glycosidases)}.
.4 {\textbf{EC:3.2.1.2} (hydrolysis of (1$\rightarrow$4)-\textalpha-D-glucosidic linkages in polysaccharides)}.
}
\medskip
%stopleaveout
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{Figures/ec_performance_by_top_level_class.png}%
  \caption{\textbf{ProteInfer performance for predictions within all 7 top-level enzyme groups.}}
  \label{fig:toplevel}
\end{figure}
Individual protein sequences can be annotated with zero (non-enzymatic proteins), one (enzymes with a single function) or many (multi-functional enzymes) leaf-level EC numbers. These are drawn from a total of 8,162 catalogued chemical reactions.
Our best $\texttt{F}_{max}$ was achieved by a model containing 5 residual blocks with 1100 filters each (full details in \textnormal{ Supplement}%htmlver:<a href="./latex/supplement.pdf">Supplement</a>
). For the dev set, $\texttt{F}_{max}$ converged within 500,000 training steps. On the random split, the model achieves $\texttt{F}_{max}$ = 0.977 (0.976-0.978) on the held-out test data. At the corresponding confidence threshold, the model correctly predicts 96.7\% of true labels, with a false positive rate of 1.4\%. Results from the clustered test set are discussed below.
Performance was roughly similar across labels at the top of the EC hierarchy, with the highest $\texttt{F}_{max}$ score observed for ligases (0.993), and the lowest for oxidoreductases (0.963) (\Cref{fig:toplevel}A%htmlver: 
). For all classes, the precision of the network was higher than the recall at the threshold maximising $\texttt{F}_{max}$. Precision and recall can be traded off against each other by adjusting the confidence threshold at which the network outputs a prediction, creating the curves shown in \Cref{fig:toplevel}B.%htmlver:the figure below
%%%INITIALANALYSIS%%%

@@

We then assessed the ability of this method to more generally localize function within a sequence, even though the model was not trained with any explicit localization information. We selected all enzymes from Swiss-Prot that had two separate leaf-node EC labels for which our model predicted known EC labels, and these labels were mappable to corresponding Pfam labels. For each of these proteins, we obtained coarse-grained functional localization by using CAM to predict the order of the domains in the sequence and compared to the ground truth Pfam domain ordering\textnormal{(see Supplement for details of the method)}%htmlver:(see <a href="./latex/supplement.pdf">Supplement</a> for details of the method)
. We found that in 296 of 304 (97\%) of the cases, we correctly predicted the ordering, though we note that the set of bifunctional enzymes for which this analysis is applicable is limited in its functional diversity \textnormal{(see Supplement)}%htmlver:(see <a href="./latex/supplement.pdf">Supplement</a>)
. Although we did not find that fine-grained, per-residue functional localization arose from our application of CAM, we found that it reliably provided coarse-grained annotation of domains' order, as supported by Pfam. This experiment suggests that this is a promising future area for research.
\subsection*{Neural networks learn a general-purpose embedding space for protein function}
Whereas InterProScan compares each sequence against more than 50,000 %from summing integrated signatures at https://www.ebi.ac.uk/interpro/release_notes.html
individual signatures and BLAST compares against an even larger sequence database, ProteInfer uses a single deep model to extract features from sequences that directly predict protein function. One convenient property of this approach is that in the penultimate layer of the network each protein is expressed as a single point in a high-dimensional space. To investigate to what extent this space is useful in examining enzymatic function, we used the ProteInfer EC model trained on the random split to embed each test set protein sequence into a 1100-dimensional vector.

@@

To investigate this, we took the subset of proteins in Swiss-Prot that are non-enzymes, and so lack any EC annotations. The network would achieve perfect accuracy on these examples if it e.g. mapped all of them to a single embedding that corresponds to zero predicted probability for every enzymatic label. Do these proteins therefore share the same representation in embedding space? The UMAP projection of these sequences' embeddings revealed clear structure to the embedding space, which we visualised by highlighting several GO annotations which the network was never supervised on. For example, one region of the embedding space contained ribosomal proteins, while other regions could be identified containing nucleotide binding proteins, or membrane proteins (\Cref{fig:unsupervised_embedding}%htmlver:interactive figure below
).
%%%nonenzymes%%%
To quantitatively measure whether these embeddings capture the function of non-enzyme proteins, we trained a simple random forest classification model that used these embeddings to predict whether a protein was annotated with the \textit{intrinsic component of membrane} GO term. We trained on a small set of non-enzymes containing 518 membrane proteins, and evaluated on the rest of the examples. This simple model achieved a precision of 97\% and recall of 60\% for an F1 score of 0.74. Model training and data-labelling took around 15 seconds. This demonstrates the power of embeddings to simplify other studies with limited labeled data, as has been observed in recent work \citep{Alley589333,biswas2020low}.
%%%SWEEP_embedding%%%

@@

An attractive property of deep learning models is that they can be run efficiently, using consumer graphics cards for acceleration. Indeed, recently, a framework has been developed to allow models developed in TensorFlow to be run locally using simply a user's browser \citep{tfjs}, but to our knowledge this has never been deployed to investigate biological sequence data. We therefore built a tool to allow near-instantaneous prediction of protein functional properties in the browser. When the user loads the tool, lightweight EC (5MB) and GO model (7MB) prediction models are downloaded and all predictions are then performed locally, with query sequences never leaving the user's computer. Inference in the browser for a 1500 amino-acid sequence takes < 1.5 seconds for both models (\Cref{sup:js_timing}%htmlver:see <a href="./latex/supplement.pdf">Supplement</a>
).
% \begin{table*}[b]
% \centering
% \caption{Bla}
% \begin{tabular}{ccccc}
% \hline
% Input               & Processor & Statistic                                 & InterProScan & ProteInfer  \\ \hline
% $1\times$ 500 aa sequence  & CPU       & Inference time   & 1 min 18 secs            & x           \\ 
% $1\times$ 500 aa sequence  & GPU       & Inference time   & \xmark️          & x           \\ 
% $ (test set) \times$ 500 aa sequences & 12x CPUs       & Inference time & 12 hours, 40 minutes            & x           \\ 
% $64\times$ 500 aa sequence & GPU       & Inference time & \xmark️          & x           \\ 
% --                  & --        & Model sizes (compressed)                  & 50 GB        & 50 MB       \\ 
% --                  & --        & Deliverable within the user’s browser     & \xmark️           &\cmark \\ \hline
% \end{tabular}
% \end{table*}
% \note{
% %I tried making "chimaeric proteins" - when N terminus of amylase is added to MAPK7 that is predicted to have extracellular localisation (otherwise unchanged). With reverse, amylase functionality is unchanged but location changes to intracellular.
% }
\subsection*{Comparison to experimental data}
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{Figures/price.png}%
  \caption{\textbf{ProteInfer predictions for a set of genes recently experimentally reannotated by high-throughput phenotyping.} ProteInfer makes confident and largely accurate predictions at the earliest levels of the EC hierarchy. Accuracy falls at the finest levels of classification (for this set of challenging genes) but fortunately the network declines to make a prediction in most cases, with every label failing to meet the threshold for positive classification. \label{fig:levels}}
\end{figure}
\begin{figure*}[b]
\centering
  \includegraphics[width=\textwidth]{Figures/tetrahydrofolate_synthase.png}
  % See https://docs.google.com/presentation/d/1PqP-Ux7far9NMYvroeZIQQ9iXi41xira7-awDC4w_ss/edit#slide=id.p for the provenance of this figure
  \caption{\textbf{Linking sequence regions to function with class activation mapping for C-1-tetrahydrofolate synthase (accession P11586).} A. Ground truth annotation of function on Uniprot \citep{uniprotlatest}. B. The three horizontal bars are the sequence region ProteInfer predicts are most involved in each corresponding reaction. This concurs with the known function localization. }
  \label{fig:cam}
\end{figure*}

@@

%startleaveout
\section*{Bibliography}
\bibliography{../bibliography}
%\onecolumn
%\newpage
%stopleaveout
%startleaveout

@@

%stopleaveout
%%%BOTTOMSTUFF%%%
%startleaveout

%stopleaveout
